\bf Supplement to "Mechanistic models for panel data: analysis of ecological experiments with four interacting species"

### 2024-04-17 ms.Rnw
UPdate

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage[authoryear,numbers]{natbib}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{hyperref}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{hyphenat}
\tolerance=1000
\usepackage{url}
\usepackage{textcomp}
\usepackage{tikz}
\usepackage{hyphenat}
\usepackage{Sweave}
\bibliographystyle{apalike}
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

@@

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Ecological experiments contain many elements or processes that are difficult to observe, so models that can describe both observable and unobservable processes are needed to explain the entire ecological experiment. As mathematical descriptions of the elements that make up a system, mechanistic models can describe the interactions between the elements of the system and their interactions with the environment. Using established processes, they can elucidate the role of hidden mechanisms \citep{Duarte2003}. The purpose of mechanistic modeling in ecology is to use mathematical models to describe the development and interactions of dynamic systems. This quantitative tool allows researchers to assess the explanatory potential of various conceptual models and to measure the magnitude of relevant parameters. Mechanistic modeling has been widely used in the social sciences, industrial engineering and physics. Now, increased computational power has made it even more effective. Because of the non-smooth, non-linear and stochastic nature of ecological information, traditional time-series analysis offers limited explanatory power. Despite creating theoretical models for the data, linking the theory to the data remains a challenging task. As a result, eco-mechanical modeling is unable to address most problems on a large scale, but only a limited range of problems \citep{levin1992problem}.\\

@@

The SIRPF model describes the dynamics of the Susceptible($S$), Infected($I$), Recovered, Food($F$), and Parasite($P$) populations as the Figure 2 and model show:

\usetikzlibrary{positioning}
\usetikzlibrary {arrows.meta}
\usetikzlibrary{shapes.geometric}

\begin{figure}
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.5cm, minimum height=0.5cm, rounded corners=.1cm, fill=blue!8},
  rhombus/.style={diamond, draw=black, minimum width=0.05cm, minimum height=0.05cm, fill=purple!8,aspect = 1,inner sep=0cm},
  travel/.style={circle, draw=black, minimum width=0.5cm, minimum height=0.5cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.6cm,inner sep=0cm},
  bendy/.style={bend left=10},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

\node (S) at (-0.5,-0.25) [square] {$S$};
\node (I) at (4.5,-0.25) [square] {$I$};
\node (P) at (2.5,0.75)  [rhombus] {$P$};
\node (R) at (3.5,2)[travel] {$R$};

\node (F) at (4.5,4.5) [report] {$F$};

\draw [->] (F) --  (R);
\draw [->] (P) --  (R);
\draw [->] (S) -- (R);
\draw [->] (I) -- (R);

\draw [->] (F) -- (S);
\draw [->] (F) -- (I);

\draw (S) edge[loop left] (S);
\draw (F) edge[loop above] (F);

\draw [->, bendy] (S) to  (I);
\draw [->, bendy] (I) to  (S);
\draw [->, bendy] (P) to (I);
% \path [->] (S) edge (I)
%            (P) edge (I);
\draw [->, bendy] (I) to (P);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. For \textit{Daphnia} either dentifera or lumholtzi, susceptible, infected \textit{Daphnia} and food retain the ability to produce offspring, which is represented in the graph as a self-recycling of $S$ and $F$, as well as transaction from $I$ to $S$. Some susceptible \textit{Daphnia} get infected by parasites, transitioning from $S$ to $I$. Both susceptible and infected \textit{Daphnia} contribute to the $R$ state upon death, releasing surviving parasites back into the environment. And the parasites continue to impact other individuals and get into another cycle. Food is consumed by both $S$ and $I$ individuals. Over time, elements in $F$ and $P$ also transit to the $R$ state.}
\label{fig:flow_SIRPF}
\end{figure}

@@

The adaptation of the SIRPF model to SF and SF2 as following allows for a more focused examination of the roles that resources play in ecological systems, especially when considering the competition or coexistence between different species. By isolating the food variable in the absence of parasites, these models can provide clearer insights into how food resource availability or scarcity impacts the population dynamics of native and invasive species.

\begin{figure}[H]
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=1cm, minimum height=1cm, rounded corners=.1cm, fill=blue!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=1cm,inner sep=0cm},
  rhombus/.style={diamond, draw=black, minimum width=0.1cm, minimum height=0.1cm, fill=purple!8,aspect = 1},
  bendy/.style={bend left=10},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

\node (Sn) at (-0.5,0) [square] {$S_n$};
\node (F) at (2.5,0) [report] {$F$};
\node (Si) at (5.5,0) [square] {$S_i$};
\node (R) at (2.5,2) [square] {$R$};

\draw [->] (F) --  (Sn);
\draw [->] (F) --  (Si);
\draw [->] (F) --  (R);
\draw [->] (Sn) --  (R);
\draw [->] (Si) --  (R);
\draw (Sn) edge[loop above] (Sn);
\draw (Si) edge[loop above] (Si);
\draw (F) edge[loop below] (F);

\draw [dotted, thick, red] (3.5, -0.75) rectangle (6.5, 1.5);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SF and SF2 model. In this system, both types of \textit{Daphnia} and the food resource contribute to reproduction, depicted as self-recycling loops for $S_n$, $S_i$, and $F$. As time progresses, individuals in $F$, $S_n$, and $S_i$ transition to the $R$ state, signifying mortality. While the whole graph show SF2 model, if we remove the node in the dashed rectangle, the graph will represent the SF model.}
\label{fig:SF&SF2}
\end{figure}

\begin{align}
&d S_n(t)=r_n S_n(t) dt- \theta_{S_n}S_n(t) dt -\delta S_n(t)dt + S_n(t) d\zeta_{S_n} \\
&d S_i(t)=r_i S_i(t) dt- \theta_{S_i} S_i(t) dt -\delta S_i(t)dt + S_i(t) d\zeta_{S_i} \\
\begin{split}
&dF(t)=\alpha F(t)\left(1-\frac{F(t)}{K_F}\right)\cdot dt-\gamma_n F(t) S_n(t) \cdot dt - \gamma_i F(t) S_i(t) \cdot dt \\
& \quad \quad \quad \quad -\delta F(t) \cdot dt+ \mu \cdot dt +F(t) d \zeta_F\\
\end{split}
\\
&\zeta_F \sim N \left( 0, \sigma_F^2dt\right) \\
&\zeta_{S_n} \sim N \left( 0, \sigma_{S_n}^2dt  \right) \\
&\zeta_{S_i} \sim N \left( 0, \sigma_{S_i}^2dt \right)
\end{align}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage[authoryear,numbers]{natbib}
\usepackage{graphicx}
\usepackage[english]{babel}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{hyperref}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{hyphenat}
\tolerance=1000
\usepackage{url}
\usepackage{textcomp}
\usepackage{tikz}
\usepackage{hyphenat}
\usepackage{Sweave}
\bibliographystyle{apalike}
\newcommand{\blind}{1}
\usetikzlibrary{positioning}
\usetikzlibrary {arrows.meta}
\usetikzlibrary{shapes.geometric}
% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

@@

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}
Ecological experiments contain many elements or processes that are difficult to observe, models that can describe both observable and unobservable processes to explain the whole ecological experiment are necessary. As a mathematical description of the elements forming a system, mechanistic modeling can describe the mutual interactions between elements in the system and their interactions with the environment \citep{Stalidzans2020}. Utilizing the established process, they elucidate the actions of hidden mechanisms  The purpose of mechanistic modeling in ecology involves employing mathematical models to depict the development and interaction of dynamic systems. This quantitative tool allows researchers to evaluate the explanatory potential of various conceptual models and measure the magnitude of pertinent parameters. Mechanistic modeling has been widely employed in social sciences, industrial engineering, and physics for an extended period. Improved computational capabilities now make it more effective. Traditional time series analysis offers limited explanatory power due to the nonstationary, nonlinear, and stochastic nature of ecological information. Despite the creation of theoretical models for the data, connecting theories with data continues to be a challenging task. As a result, ecological mechanical modeling cannot solve most of the problems on a wide scale, but rather, only a limited range of the problems \\

@@

@@

The adaptation of the SIRPF model to SF and SF2 as following allows for a more focused examination of the roles that resources play in ecological systems, especially when considering the competition or coexistence between different species. By isolating the food variable in the absence of parasites, these models can provide clearer insights into how food resource availability or scarcity impacts the population dynamics of native and invasive species.
\begin{align}
&d S_n(t)=r_n S_n(t) dt- \theta_{S_n}S_n(t) dt -\delta S_n(t)dt + S_n(t) d\zeta_{S_n} \\
&d S_i(t)=r_i S_i(t) dt- \theta_{S_i} S_i(t) dt -\delta S_i(t)dt + S_i(t) d\zeta_{S_i} \\
\begin{split}
&dF(t)=\alpha F(t)\left(1-\frac{F(t)}{K_F}\right)\cdot dt-\gamma_n F(t) S_n(t) \cdot dt - \gamma_i F(t) S_i(t) \cdot dt \\
& \quad \quad \quad \quad -\delta F(t) \cdot dt+ \mu \cdot dt +F(t) d \zeta_F\\
\end{split}
\\
&\zeta_F \sim N \left( 0, \sigma_F^2dt\right) \\
&\zeta_{S_n} \sim N \left( 0, \sigma_{S_n}^2dt  \right) \\
&\zeta_{S_i} \sim N \left( 0, \sigma_{S_i}^2dt \right)
\end{align}

### 2024-04-17 si/si.Rnw
UPdate

\section{Model Diagram}

Dynamic models addressed in this manuscript are comprehensively delineated through the descriptions provided within the main text, augmented by the supplementary details furnished in this section. Notwithstanding their complete mathematical specification, visual diagrams of dynamic systems frequently facilitate comprehension of the underlying equations. Hence, this section presents a model diagram, which corresponds to the SF model. It is imperative to recognize that, given the models are articulated through their mathematical formulations and numerical implementations, these diagrams do not constitute unique visual depictions. There exist alternative visualizations that might further aid in elucidating the models discussed herein.

>>>>

\section{Model Diagram}
\subsection{SIRPF Model}

The SIRPF model describes the dynamics of the Susceptible($S$), Infected($I$), Recovered, Food($F$), and Parasite($P$) populations as the Figure 2 and model show:

\usetikzlibrary{positioning}
\usetikzlibrary {arrows.meta}
\usetikzlibrary{shapes.geometric}

\begin{figure}
\begin{center}
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.5cm, minimum height=0.5cm, rounded corners=.1cm, fill=blue!8},
  rhombus/.style={diamond, draw=black, minimum width=0.05cm, minimum height=0.05cm, fill=purple!8,aspect = 1,inner sep=0cm},
  travel/.style={circle, draw=black, minimum width=0.5cm, minimum height=0.5cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.6cm,inner sep=0cm},
  bendy/.style={bend left=10},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

\node (S) at (-0.5,-0.25) [square] {$S$};
\node (I) at (4.5,-0.25) [square] {$I$};
\node (P) at (2.5,0.75)  [rhombus] {$P$};
\node (R) at (3.5,2)[travel] {$R$};

\node (F) at (4.5,4.5) [report] {$F$};

\draw [->] (F) --  (R);
\draw [->] (P) --  (R);
\draw [->] (S) -- (R);
\draw [->] (I) -- (R);

\draw [->] (F) -- (S);
\draw [->] (F) -- (I);

\draw (S) edge[loop left] (S);
\draw (F) edge[loop above] (F);

\draw [->, bendy] (S) to  (I);
\draw [->, bendy] (I) to  (S);
\draw [->, bendy] (P) to (I);
\draw [->, bendy] (I) to (P);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SIRPF model. Apart from the quantified states above, the graph includes $R$ state, which represents deadth status. For \textit{Daphnia} either dentifera or lumholtzi, susceptible, infected \textit{Daphnia} and food retain the ability to produce offspring, which is represented in the graph as a self-recycling of $S$ and $F$, as well as transaction from $I$ to $S$. Some susceptible \textit{Daphnia} get infected by parasites, transitioning from $S$ to $I$. Both susceptible and infected \textit{Daphnia} contribute to the $R$ state upon death, releasing surviving parasites back into the environment. And the parasites continue to impact other individuals and get into another cycle. Food is consumed by both $S$ and $I$ individuals. Over time, elements in $F$ and $P$ also transit to the $R$ state.}
\label{fig:flow_SIRPF}
\end{figure}

\subsection{SF2 Model}

\begin{figure}[H]
\begin{center}
%%%%% SEAIR diagram
\resizebox{8cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=1cm, minimum height=1cm, rounded corners=.1cm, fill=blue!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=1cm,inner sep=0cm},
  rhombus/.style={diamond, draw=black, minimum width=0.1cm, minimum height=0.1cm, fill=purple!8,aspect = 1},
  bendy/.style={bend left=10},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

\node (Sn) at (-0.5,0) [square] {$S_n$};
\node (F) at (2.5,0) [report] {$F$};
\node (Si) at (5.5,0) [square] {$S_i$};
\node (R) at (2.5,2) [square] {$R$};

\draw [->] (F) --  (Sn);
\draw [->] (F) --  (Si);
\draw [->] (F) --  (R);
\draw [->] (Sn) --  (R);
\draw [->] (Si) --  (R);
\draw (Sn) edge[loop above] (Sn);
\draw (Si) edge[loop above] (Si);
\draw (F) edge[loop below] (F);

\draw [dotted, thick, red] (3.5, -0.75) rectangle (6.5, 1.5);
\end{tikzpicture}
}
\end{center}
\vspace{-5mm}
\caption{A flow diagram for the SF and SF2 model. In this system, both types of \textit{Daphnia} and the food resource contribute to reproduction, depicted as self-recycling loops for $S_n$, $S_i$, and $F$. As time progresses, individuals in $F$, $S_n$, and $S_i$ transition to the $R$ state, signifying mortality. While the whole graph show SF2 model, if we remove the node in the dashed rectangle, the graph will represent the SF model.}
\label{fig:SF&SF2}
\end{figure}

\subsection{SF Model}
Dynamic models addressed in this manuscript are comprehensively delineated through the descriptions provided within the main text, augmented by the supplementary details furnished in this section. Notwithstanding their complete mathematical specification, visual diagrams of dynamic systems frequently facilitate comprehension of the underlying equations. Hence, this section presents a model diagram, which corresponds to the SF model. It is imperative to recognize that, given the models are articulated through their mathematical formulations and numerical implementations, these diagrams do not constitute unique visual depictions. There exist alternative visualizations that might further aid in elucidating the models discussed herein.
